---
title: "Working with Tabular Data"
author: "Gupta, Rohan"
date: "May 13th, 2025"
output:
  html_document:
    df_print: paged
subtitle: "Analysis of Wines"
---

## Question 1

I created a new project for this .Rmd file, formatted it, and added my work to my private github repo.


## Question 2

```{r q2_loadCSV, echo=FALSE}
# Start by loading the data using the Canvas URL
df.wines <- read.csv("https://s3.us-east-2.amazonaws.com/artificium.us/datasets/whitewines.csv",
                 stringsAsFactors = FALSE)
```

After reading through the text that needs to be reproduced, I mapped out the calculations I needed to conduct:

-   Total number of rows/observations
-   Percent of wines with alcohol content over 11%
-   Max alcohol content 
-   Scatter plot showing sugar vs alcohol content

I've completed the above calculations and assigned values to variables in the below code chunk.

```{r q2_calculcations, echo=FALSE}
# Total number of rows
total_wines <- nrow(df.wines)

# Percent of wines with alc > 11
wines_above_11 <- sum(df.wines$alcohol > 11)/total_wines
wines_above_11 <- round(wines_above_11 * 100, 1)

# Max alc content 
max_alc <- max(df.wines$alcohol)
```


After completing all the necessary calculations, I reproduced the formatted text and plot from Canvas below: 



### **Analysis of Wines**

There are a total of <mark>`r total_wines`</mark> wines in this data set. <mark>`r wines_above_11`</mark>% of the wines have an alcohol content over 11%. the highest alcohol content of any wine is <mark>`r max_alc`</mark>%. 

The scatter plot below illustrates the correlation between alcohol and sugar content. 

```{r q2_plot, echo=FALSE}
# Plot sugar against alc content
plot(x = df.wines$residual.sugar,
     y = df.wines$alcohol,
     xlab = "Sugar",
     ylab = "Alcohol",
     main = "Sugar vs Alcohol Content",
     col.main = "darkblue",
     type = "p")
```

