---
title: 'ASSIGNMENT 06.2: Query Tabular Data with SQL'
author: "Rohan Gupta"
date: "June 18th, 2025"
output:
  html_document:
    df_print: paged
---


[Reference for Setting Default Chunk Options](https://rmarkdown.rstudio.com/lesson-3.html)

```{r setup, echo=FALSE}
# Load knitr for default chunk settings and kable
library(knitr)
# Set echo to false and supress warnings
opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r libraries}
# Load kableExtra for styling
library(kableExtra)
# Load the sqldf package
library(sqldf)
```


## Analysis of Patient Data


```{r loadData}
# For the sake of keeping the knitted file clean and only relevant to the format
# in the assignmetn example, all my code explanations are within chunks as
# comments. Chunk names describe which part of the analysis each chunk is
# responsible for.



# Start by loading in the data
df <- read.csv("https://s3.us-east-2.amazonaws.com/artificium.us/datasets/diabetes_data.csv",
               header = TRUE,
               stringsAsFactors = FALSE)
```


```{r totalRecords}
# Create the total patient records query; I'm simply getting a row count of the
# entire df
ptRecsQ <- "
SELECT 
  COUNT(*) AS TotalPtRecs
FROM 'df'
"

# Pass the query to sqldf
ptRecs <- sqldf(ptRecsQ)

# Get the val from the resulting df 
ptRecsVal <- ptRecs$TotalPtRecs[1]
```


```{r meanAge}
# Create queries getting the average age for the entire dataset, then with a
# WHERE clause filtering by sex
overallMeanQ <- "
SELECT
  AVG(Age) AS OverallAvgAge
FROM 'df'
"

maleMeanQ <- "
SELECT
  AVG(Age) AS MaleAvgAge
FROM 'df'
WHERE Sex = 0
" 

femaleMeanQ <- "
SELECT
  AVG(Age) AS FemaleAvgAge
FROM 'df'
WHERE Sex = 1
" 

# Execute each query; these return as single col single row dfs (same as the
# first query)
overallMean <- sqldf(overallMeanQ)

maleMean <- sqldf(maleMeanQ)

femaleMean <- sqldf(femaleMeanQ)


# Get the value from each of the dataframes
overallMeanVal <- overallMean$OverallAvgAge[1]
maleMeanVal <- maleMean$MaleAvgAge[1]
femaleMeanVal <- femaleMean$FemaleAvgAge[1]

```


```{r fruitsAndVeg}
# Here, I'm querying to just get the number of patients that eat both fruits and
# veggies; in the markdown, I'll divide this by the total patient records value
fruitsVegQ <- "
SELECT
  COUNT(*) AS FruitsAndVegCount
FROM 'df'
WHERE
  Fruits = 1
  AND Veggies = 1
"

# Execute the query
fruitsVeg <- sqldf(fruitsVegQ)

# Get the val
fruitsVegVal <- fruitsVeg$FruitsAndVegCount[1]

```


```{r smokers}
# Here, I'm getting the smokers count, which I'll also divide bu the total
# patients in the markdown
smokersQ <- "
SELECT
  COUNT(*) AS SmokerCt
FROM 'df'
WHERE Smoker = 1
"

# Execute the query
smokers <- sqldf(smokersQ)

# Get the val
smokersVal <- smokers$SmokerCt[1]

```


```{r bmiOver26}
# Query to get the number of pts with a BMI > 26 and no phys activity
bmiQ <- "
SELECT
  COUNT(*) AS BmiNoPA
FROM 'df'
WHERE
  BMI > 26
  AND PhysHlth = 0
"

bmi <- sqldf(bmiQ)

bmiVal <- bmi$BmiNoPA[1]

```



The data set has <mark>`r ptRecsVal`</mark> patient records. Males were coded in the data as 0 while females were coded as 1. The mean age for all patients was <mark>`r round(overallMeanVal, 2)`</mark> years (female <mark>`r round(femaleMeanVal, 2)`</mark>, male <mark>`r round(maleMeanVal, 2)`</mark>). <mark>`r signif((fruitsVegVal / ptRecsVal) * 100, 3)`</mark>% of all patients reported eating both fruits and vegetables. <mark>`r signif((smokersVal/ptRecsVal) * 100, 3)`</mark>% of all patients were smokers. <mark>`r signif((bmiVal / ptRecsVal) * 100, 3)`</mark>% of all patients with a BMI over 26 did not report any physical activity.



## Summary of Health Characteristics

```{r summaryTableMale}
# To create this summary table, I've decided to use sqldf() queries to first
# retrieve each piece of data I need, then combine everything into an r
# data.frame. 

# I already have a queried value for average age (for both genders). For the
# following queries, I'll get the value for each constraint (for example, the
# actual number value of pts diagnosed with diabetes) and divide this value by
# the total number of pts inside the final df.


# In this chunk, I'll be starting with the metrics for male patients in the df;
# I've set the names of the cols to the name of the vars so I can select use the
# same string for subsetting in the next chunk


# Diabetes query, df, and value
diabQ <- "
SELECT 
  COUNT(*) AS diabQ
FROM 'df'
WHERE 
  Diabetes = 1
"


# High chol query, df and value
cholQ <- "
SELECT 
  COUNT(*) AS cholQ
FROM 'df'
WHERE HighChol = 1
"


# Here, I use a subquery to filter the rows where BMI is greater than 2 SDs +
# BMI mean; this subquery becomes a single value used to filter using WHERE, and
# the number of rows where this condition is satisfied is returned by COUNT(*).
# I'll also note that this is basically the rows where z-score = 2, so that's
# what I've named the row in the final df
bmiSdQ <- "
SELECT
  COUNT(*) AS bmiSdQ
FROM 'df'
WHERE BMI > (
  SELECT 
    AVG(BMI) + 2 * STDEV(BMI)
  FROM 'df'
)
"


# No physical activity
noPhysQ <- "
SELECT 
  COUNT(*) AS noPhysQ
FROM 'df'
WHERE PhysHlth = 0
"


# High bp
highBpQ <- "
SELECT 
  COUNT(*) AS highBpQ
FROM 'df'
WHERE HighBP = 1
"
```


```{r maleVals}
# Create a named list of the character string queries from above
queries <- list(diabQ = diabQ, 
                cholQ = cholQ, 
                bmiSdQ = bmiSdQ, 
                noPhysQ = noPhysQ, 
                highBpQ = highBpQ)

# Initialize an empty list where I'll store the male queries
maleQueries <- list()

# Loop over the generic queries
for (q in 1:length(queries)) {
  
  # For each generic query, I'll add the condition to filter for just male pts
  maleQueries[[q]] <- paste0(queries[[q]], " AND Sex = 0")
  
  # Here, I extract the variable name for each element in the named list (which
  # will serve as the base of the new variable name for the new 'male' query)
  qName <- names(queries)[q]
  
  # Assign a var for the output of the male query being passed to sqldf()
  qExecute <- sqldf(maleQueries[[q]])
  
  # Subset to extract the value from the single value dfs (col = qName, row = 1)
  qVal <- qExecute[, qName][1]
  
  # And finally assign a var name (the generic query var name + "M" for male) to
  # the output of the above male query
  assign(paste0(qName, "M"), qVal)

  
  # This loop returns 5 new variables, all df's containing the result for each
  # generic query for male patients
}

```



```{r femaleVals}
# Here, I'm using the same generic queries and loop format as above, but
# changing the added condition for each query to filter for female pts. I'm also
# assigning the results to output vars called genericVarName + "F"

# Initailize an empty list to store female queries
femaleQueries <- list()


# Same loop, just shortened comments
for (q in 1:length(queries)) {
  # Change the condition to filter for Female pts
  femaleQueries[[q]] <- paste0(queries[[q]], " AND Sex = 1")
  
  # Get the generic q name
  qName <- names(queries)[q]
  
  # Get the output of the female query as a df
  qExecute <- sqldf(femaleQueries[[q]])
  
  # Subset by col name (same as query name) to get the first row val
  qVal <- qExecute[, qName][1]
  
  # Assign the value to queryF var
  assign(paste0(qName, "F"), qVal)
}

```



```{r createFinalDf}
# Here, I'm initializing the empty final table
finalDf <- data.frame(
  Male = numeric(6),
  Female = numeric(6)
)

# Assigning row names for the df
rownames(finalDf) <- c("Average Age",
                       "Diabetes (%)",
                       "High Cholesterol (%)",
                       "ZScore > 2 (%)",
                       "No Physical Activity (%)",
                       "High Blood Pressure (%)")



# Get the male and female value variable names as strings (doing this manually
# since I already created the vars with assign())
mValNames <- paste0(names(queries), "M")
fValNames <- paste0(names(queries), "F")

# Add the mean age for male and female to their respective vectors (so I can
# easily pass everything in order to the df in the next loop)
mValAll <- c("maleMeanVal", mValNames)
fValAll <- c("femaleMeanVal", fValNames)

# Loop over the vars from 
for (row in 1:nrow(finalDf)) {
  # Populate the first col (Male) with the values from the male extracted values
  # using their variable names; each row index already corresponds to the index
  # of values in ValNames
    # From reference: using get() to get the value from the string var
  finalDf[row, 1] <- get(mValAll[row])
  
  # Same for female values
  finalDf[row, 2] <- get(fValAll[row])
  
}

# Format the final table with rounding
# Age to 2 significant digits
finalDf[1, ] <- signif(finalDf[1, ], 3)
# Divide each val by total pt count, round to 2 decimals
finalDf[2:6, ] <- round((finalDf[2:6, ] / ptRecsVal) * 100, 2)


# Load just to run pipe operator below
library(magrittr)

# Display the table using kable (from kable docs)
finalDf %>%
  kbl(caption = "Patient Metrics by Gender") %>%
  kable_classic(c("striped", "hover")) %>% 
  kable_styling(font_size = 20, full_width = TRUE)

```







*"maleVals" Chunk References:*

[Reference to Get Var Names using a Named List](https://stackoverflow.com/questions/42227416/extract-variable-names-from-list-or-vector-in-r)

[Reference for Getting Single Element Name from Named List](https://stackoverflow.com/questions/9048375/extract-names-of-objects-from-list)

[Reference for Dynamic Variable Names in a Loop Using assign()](https://stackoverflow.com/questions/16566799/change-variable-name-in-for-loop-using-r)


*"createFinalDf" Chunk References:*

[Reference for Setting DF Row Names](https://stat.ethz.ch/R-manual/R-devel/library/base/html/row.names.html)

[Reference for Getting Values from String Vars with get()](https://stackoverflow.com/questions/3971844/access-variable-value-where-the-name-of-variable-is-stored-in-a-string)

[Reference for Adding Value to the Start of a Vector](https://stackoverflow.com/questions/60762376/adding-element-to-beginning-of-the-vector-in-r)



